Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-05-16 Text/migration.tex
Addressed many concerns and A_m(t) plot. Still waiting on a bit of periodogram analysis.
	new file:   Text/Figures/amt.eps
	modified:   Text/migration.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@76 3b7f7548-5c66-0410-83ab-3997167635fc

{\bf See note in Latex file.}
%Scott: Verify the accuracy of all this.

The model disk, based on an equlibrium model in \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of ??? at radial grid zone ??? (??? AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density is given an initial 0.01 \% random cell-to-cell perturbation.

@@

%Scott: A number of important assertions in this paragraph need to verified by further analysis.
{\it The $t = 0$ Planet Run.} {\bf See note in Latex file.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.8 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at ??? AU. Modes of cos($m\phi$) symmetry with $m =$ 5 and 6 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORPs earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. Beginning at 3 ORP, when the $m = 3$ mode has grown significantly and the planet is lagging one of its arms in azimuth, the planet gains angular momentum and moves outward. At about $t =$ 4 ORP, as the planet overtakes the arm, the sign of the net torque reverses and the planet plunges from 23 to 17 AU in about 4 ORP (10$^3$ yr). During this time, the planet and mode seem to interact strongly enough that the planet essentially surfs along the arm, maintaining a location within the arm as it plunges. After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values are comparably strong. The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at ??? 28 ??? AU in the GI-active disk in this early part of the asymptotic phase.

%Scott: Even more work is needed to verify (or refute) what is said here. Some of what I wrote is pure speculation and needs to be checked.
{\it The $t = 10$ Planet Run.} {\bf See note in Latex file.} The motion of the planet in this case is more difficult to interpret. We see intervals of fairly rapid inward or outward migration over several orbits, as well as times when radial migration appears to stall. Between $t = 15$ and 19 ORP, the pattern of outward migration for 2 ORP followed by an inward plunge over the next 2 ORP resembles the behavior in the $t = 0$ planet run during the burst, but, in this case, there is no distinct phase of GI activity. Animations of the simulation show there is a complex interaction between the planet, its spiral wake, and the global spiral arms caused by the instabilities. Analysis of images and periodicities suggests that, between 15 and 19 ORP, the planet may have passed through one arm of an $m =$ ??? pattern with CR at ??? AU. It is possible that the planet actually stimulated this particular mode, in manner similar to what occurred in the $t = 0$ planet run, because this $m =$ ??? periodicity is absent over a similar time interval in the fiducial run.

@@

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities (Li and Lovelace articles abound 2000, 2001). Meschiari \& Laughlin (2008) demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs. {\bf Scott Comment: say more about gap opening criteria and whether they apply here. Dick's reply in tex file.}
%Scott: Do it. I am not sure what you have in mind. Remember, however, that we are already at about maximum length for Letter.

%Scott and Aaron: Can you go over this part carefully and fill in some calculation for the migration rate?
Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae that are fully applicable to our radiatively cooled disk, using equation ??? from ??? with parameters from the simulated disk, we estimate ??? AU yr$^{-1}$ as an analytic expectation. This is probably a reasonable estimate for comparison purposes, because our resolution is probably insufficient to compute the corotation torque \citep{paardekooper2010}. 
%Scott and Aaron: Is this right?
While the interaction with a GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in the GI-active disk is highly variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In the case of the $t = 10$ planet run, the overall radial migration rate is similar but even more erratic in direction. The important departures from laminar disk theory are that there are times when migration stalls and that interactions with discrete low-order modes seem to be important. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

%Scott and Aaron: This is REALLY speculative. We need analysis here! The following is a modified version of text from Scott's poster mixed with the idea from my dream.
In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final radial positions of both planets happens to be the location of the Inner Lindblad Resonance (ILR) for the strong m = 2 mode present in the fiducial run. This ILR is associated with a surface density enhancement in the disk. 
%Is this true? As the planet reaches the outer edge of a surface density enhancement, the local density gradient may be so steep that the ILR torques of the planet cancel the OLR torques of the planet. This would probably put the planet right outside the ILR of a major mode, because this is where we expect pile ups of material. (This came to me in a dream.) We can actually measure the surface density gradient near 17 AU and test this directly. 
As the planet approaches this enhancement, the surface density gradient may be falling steeply enough to halt migration, a mechanism can work in a laminar disk. 
%The words that follow are from Scott. We must DO the analyses of periods, look at the surface density, look at the Q distribution (I doubt the planets are out of the GI active region yet), and think about the physics.
Alternatively, at 17 AU, the level of GI activity is quite different interior to the planet ($Q \approx$ ???) and exterior to the planet ($Q \approx$ ???), and this may affect the balance of the Lindblad torques exerted by the disk on the planet. Whether the migration really does stall and what the mechanism may be are subjects for further research.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}

>>>>

The model disk, based on an equlibrium model in \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density is given an initial 0.01 \% random cell-to-cell perturbation.

@@

%Scott: A number of important assertions in this paragraph need to verified by further analysis.
{\it The $t = 0$ Planet Run.} {\bf See note in Latex file.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.8 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ 2AU. Modes of cos($m\phi$) symmetry with $m =$ 5 and 6 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORPs earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. Beginning at 3 ORP, when the $m = 3$ mode has grown significantly and the planet is lagging one of its arms in azimuth, the planet gains angular momentum and moves outward. At about $t =$ 4 ORP, as the planet overtakes the arm, the sign of the net torque reverses and the planet plunges from 23 to 17 AU in about 4 ORP (10$^3$ yr). During this time, the planet and mode seem to interact strongly enough that the planet essentially surfs along the arm, maintaining a location within the arm as it plunges. After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values are comparably strong. The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at 29 AU in the GI-active disk in this early part of the asymptotic phase.

%Scott: Even more work is needed to verify (or refute) what is said here. Some of what I wrote is pure speculation and needs to be checked.
{\it The $t = 10$ Planet Run.} {\bf See note in Latex file.} The motion of the planet in this case is more difficult to interpret. We see intervals of fairly rapid inward or outward migration over several orbits, as well as times when radial migration appears to stall. Between $t = 15$ and 19 ORP, the pattern of outward migration for 2 ORP followed by an inward plunge over the next 2 ORP resembles the behavior in the $t = 0$ planet run during the burst, but, in this case, there is no distinct transition between phases of GI activity. Animations of the simulation (available at {\bf BLAH}) show there is a complex interaction between the planet, its spiral wake, and the global spiral arms caused by the instabilities. Analysis of images and periodicities suggests that, between 15 and 19 ORP, the planet may be locked in a resonance with an arm of a $m = 2$ or 3 pattern both of which have CR at 27 AU. Over this time interval the average orbit period of the planet is 0.5 ORP, while both the $m = 2$ and 3 patterns have a pattern period of $\sim 0.74$ ORP, suggesting a 2:3 resonance. It is possible that the planet actually stimulated one or both of these modes, in manner similar to what occurred in the $t = 0$ planet run, because this $m =$ ??? periodicity is absent over a similar time interval in the fiducial run.

@@

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities \citep{li1999,li2000}. \citet{meschiari2008} demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not quite massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs.

%Scott and Aaron: Can you go over this part carefully and fill in some calculation for the migration rate?
Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae that are fully applicable to our radiatively cooled disk, using equation 70 from \citet{tanaka2002} with parameters from the simulated disk, we estimate 5.8$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. This is probably a reasonable estimate for comparison purposes, because our resolution is probably insufficient to compute the corotation torque \citep{paardekooper2010}. 
%Scott and Aaron: Is this right?
While the interaction with a GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in the GI-active disk is highly variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In the case of the $t = 10$ planet run, the overall radial migration rate is similar but even more erratic in direction. The important departures from laminar disk theory are that there are times when migration stalls and that interactions with discrete low-order modes seem to be important. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

%Scott and Aaron: This is REALLY speculative. We need analysis here! The following is a modified version of text from Scott's poster mixed with the idea from my dream.
In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final radial positions of both planets happens to be the location of the Inner Lindblad Resonance (ILR) for the strong m = 2 mode present in both runs. This ILR is associated with a surface density enhancement in the disk. 
%Is this true? As the planet reaches the outer edge of a surface density enhancement, the local density gradient may be so steep that the ILR torques of the planet cancel the OLR torques of the planet. This would probably put the planet right outside the ILR of a major mode, because this is where we expect pile ups of material. (This came to me in a dream.) We can actually measure the surface density gradient near 17 AU and test this directly. 
As the planet approaches this enhancement, the surface density gradient may be falling steeply enough to halt migration, a mechanism can work in a laminar disk. 
%The words that follow are from Scott. We must DO the analyses of periods, look at the surface density, look at the Q distribution (I doubt the planets are out of the GI active region yet), and think about the physics.
Alternatively, at 17 AU, the level of GI activity is quite different interior to the planet ($Q \gtrsim 2$) and exterior to the planet ($Q \lesssim 2$), and this may affect the balance of the Lindblad torques exerted by the disk on the planet. Whether the migration really does stall and what the mechanism may be are subjects for further research.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/amt.eps}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}
